Contracts for unions and intersections are hard for Nickel

### 2021-05-27 paper.tex
Pass on intro

This is why the authors have been developing the Nickel
language\cite{NickelRepo}. Nickel is a configuration programming language,
meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
lightweight and easy to write while providing support for writing correct code
and generating valid data. Nickel originated as a rethink of the language of the
Nix package manager, used to describe one of the biggest sets of packaged
software in existence\cite{repology}.  Nickel targets build systems, cloud
deployment tools (Kubernetes, Terraform or NixOps), or anything requiring a
non-trivial configuration as well.

\subsection*{To type or not to type}

@@

\begin{description}
    \item[(correctness)] The program runs without errors. Errors may be type
        errors, raised for example when evaluating the nonsensical division
        \nickel{1 / "ftp"}. More generally, they can any the violation of a pre-
        or post-condition. When such an error cannot be prevented, we want at
        least early failures and relevant error messages.
    \item[(validity)] The program generates valid data, i.e. data that
        conforms to a schema (imposed by the system consuming the
        configuration). For example, when configuring a systemd service, the
        schema may require the presence of a field \lstinline+port+ that holds a
        valid port number.
\end{description}

A traditional answer is to use static types, in particular static types for
correctness. However, Nickel programs are of a special kind: they are usually
simple, terminating programs that run on fixed inputs. Correctness errors will
show up at evaluation anyway, or will not matter (dead code). With respect to
validation, traditional types seem too limited. Validation commonly involves
checking requirements such as being a valid port number or a valid URL which
are, while not technically impossible, really hard to check statically.

In a dynamic language, it is natural and easier to validate data at runtime.
This underlies the approach of dynamic typing. But dynamically typed languages
usually only check the same simple predicates as static types (being a number,
being a string, etc.), and solely for data that are arguments to a primitive
operation. For validation, we want to check arbitrary predicates on arbitrary
values. This is provided in a basic form by assertions, available in most
languages out there. An assertion is an expression that verifies that a
predicate holds, or abort the program otherwise.  While assertions are fine for
simple sanity checks, they can quickly becomes verbose, clumsy to compose and to
reuse, and provide very limited context attached to an error. It turns out there
is a better tool than plain assertions: contracts.

@@

Assertions are usually mainly used for correctness by enforcing pre- and
post-conditions of functions at runtime. In their foundational
paper\cite{FindlerFelleisenHOContracts}, Findler and Felleisen introduce a
principled approach to run-time assertion checking called contracts. As opposed
to plain assertions, contracts can be composed naturally into more complex ones.
Nickel features such a contract system. For example, contracts nicely support
higher-order functions. If the user has written an \nickel{Even} contract for
number, they can use it right away to form a \nickel{#Even -> #Int} contract, as
illustrated in Figure~\ref{fig:contract-higher-order}. This contract will check
at each call that the argument is an even number (pre-condition) and that the
return value is a integer (post-condition). The \nickel{#} character
distinguishes contracts in Nickel. Contracts also introduce the notion of blame
which is crucial to good error reporting in presence of higher-order contracts
and other composed contracts.

@@

In our opinion, one of the main appeals of Nickel is this ability to write a
simple validation function for, say, a port number, and have it immediately
available as a contract that can be used to construct other contracts by
writing natural type-like expressions. Another example is given in
Figure~\ref{fig:contract-composition}. The first line defines a port contract
from a predicate and use it in the definition of a larger
\lstinline+Configuration+ record contract. The \lstinline+Configuration+
contract checks that a value is a record with the following shape: it must have
a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
a \emph{schema} used to validate the final data, which is exactly what we were
looking for in the previous section. Not only it plays the role of a schema, but
it also \emph{looks} like one, as opposed to a less readable plain validation
function. The \lstinline+|+ operator entails that a value respects a contract,
akin to \lstinline+:+ for types. We assume that a \lstinline+Host+ contract has
been defined in a similar way. \unsure{What's missing in this section is an
example of how contracts differ from standard dynamic type checking on some
function. Namely, the quality of error messages.}

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | #Host,
  port | #Port,
} in
{host = "nickel-lang.org", port = 80}
| #Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:contract-composition}
\end{figure}
\unsure{I wouldn't use nickel-lang.org unless you are
planning on putting something up there. Maybe nixos.org?}

Contracts have another perk: they correspond to the type casts of gradual
typing, a discipline for mixing static and dynamic typing in the same language.
As such, they are a precious ingredient for handling the interaction of
statically typed and dynamically typed code. While we focused on the writing of
configuration, Nickel is a proper programming language that is also designed to
write generic and reusable library code. There, the typing trade-off is
different, and static typing is more adapted. Nickel has a gradual type system,
which is yet another argument for having contracts. Let us end this section by
stressing the following points:
\begin{itemize}
    \item \textbf{Even in a purely dynamically typed language, contracts are an
        appealing feature for validation and correctness}. The gradual
        typing part deserved to be mentioned, but is is out of the scope of this
        paper. 
    \item \textbf{User-defined contracts are essential for validation}. Some contract
        systems (as the one of~\cite{RootCauseOfBlame} that will be extensively
        discussed later in the paper) do not feature user-defined contracts.
        Rather, a fixed set of built-in contracts such as \nickel{Number} or
        \nickel{String} are provided, very much like base types. The absence of
        user-defined contracts is understandable when using contracts as a
        dynamic typing discipline for correctness. But a system without
        user-defined contracts is limited in the same way as static typing with
        respect to validation.
\end{itemize}

>>>>

This is why the authors have been developing the Nickel
language\cite{NickelRepo}. Nickel is a configuration programming language,
meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
lightweight and easy to write while providing support for writing correct code
and generating valid data. Nickel originated as a rethink of the language of the
Nix package manager, used to describe one of the biggest sets of packaged
software in existence\cite{repology}.  Nickel targets build systems, cloud
deployment tools (Kubernetes, Terraform or NixOps), or anything requiring a
non-trivial configuration.

\subsection*{Correct programs \& valid data}

@@

\begin{description}
    \item[(correctness)] The program runs without errors. Errors may be type
        errors, raised for example when evaluating the nonsensical division
        \nickel{1 / "ftp"}. More generally, they can be any the violation of a
        pre- or post-condition. When such an error cannot be prevented, we want
        at least early failures and relevant error messages.\todo{make this
        definition more precise}
    \item[(validity)] The program generates valid data, i.e. data that conforms
        to a schema (imposed by the system consuming the configuration). For
        example, when configuring a web service, the schema may require the
        presence of a field \lstinline+port+ that holds a valid port number.
\end{description}

A traditional answer is to use static types, and in particular static types for
correctness. However, Nickel programs are of a special kind: they are usually
simple, terminating programs that run on fixed inputs. Correctness errors will
show up at evaluation anyway, or will not matter (dead code). With respect to
validation, traditional types are too limited. Validation commonly involves
checking requirements such as being a valid port number or a valid URL which
are, while not technically impossible, really hard to check statically.

In a dynamic language, it is natural and easier to validate data at runtime.
This underlies the approach of dynamic typing. But dynamically typed languages
usually only check the same simple predicates as static types (being a number,
being a string, etc.), and solely for data that are arguments to a primitive
operation. For validation, one needs to check arbitrary predicates on arbitrary
values. This is provided in a basic form by assertions, available in most
languages out there. An assertion is an expression that verifies that a
predicate holds, or abort the program otherwise.  While assertions are fine for
simple sanity checks, their use can incur verbosity, they are clumsy to compose
and to reuse, and they provide limited context attached to an error. It turns
out there is a better tool than plain assertions: contracts.

@@

Assertions are usually mainly used for correctness by enforcing pre- and
post-conditions of functions at runtime. In their foundational
paper\cite{FindlerFelleisenHOContracts}, Findler and Felleisen introduce a
principled approach to run-time assertion checking called contracts. As opposed
to plain assertions, contracts can be composed naturally into more complex ones.
Nickel features such a contract system. For example, contracts nicely support
higher-order functions. If the user has written an \nickel{Even} contract for
numbers, they can use it right away to form a function contract \nickel{#Even ->
#Int} contract, as illustrated in Figure~\ref{fig:contract-higher-order}. This
contract will check at each call that the argument is an even number
(pre-condition) and that the return value is a integer (post-condition). The
\nickel{#} character distinguishes contracts in Nickel. Contracts also introduce
the notion of blame which is crucial to good error reporting in presence of
higher-order contracts and other composed contracts.

@@

In our opinion, one of the main appeals of Nickel is this ability to write a
simple validation function for, say, a port number, and have it immediately
available to construct other contracts by writing simple type-like expressions.
Another example is given in Figure~\ref{fig:contract-composition}. The first
line defines a port contract from a predicate and use it in the definition of a
larger \lstinline+Configuration+ record contract. The \lstinline+Configuration+
contract checks that a value is a record with the following shape: it must have
a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
a \emph{schema} used to validate the final data, which is exactly what we were
looking for in the previous section. Not only it plays the role of a schema, but
it also \emph{looks} like one, as opposed to the corresponding plain validation
function. The \lstinline+|+ operator entails that a value respects a contract,
akin to \lstinline+:+ for types. We assume that a \lstinline+Host+ contract has
been defined in a similar way. \unsure{What's missing in this section is an
    example of how contracts differ from standard dynamic type checking on some
function. Namely, the quality of error messages.}

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | #Host,
  port | #Port,
} in
{host = "nixos.org", port = 80}
| #Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:contract-composition}
\end{figure}
\resolved{I wouldn't use nickel-lang.org unless you are
planning on putting something up there. Maybe nixos.org?}

Contracts have another perk: they correspond to the type casts of gradual
typing, a discipline for mixing static and dynamic typing in the same language.
As such, they are a precious ingredient for handling the interaction of
statically typed and dynamically typed code. While we focused on the writing of
configuration, Nickel is a proper programming language that is also designed to
write generic and reusable library code. There, the typing trade-off is
different, and static typing is more adapted. Nickel has a gradual type system,
which is yet another argument for having contracts. Let us end this section by
stressing the following points:
\begin{itemize}
    \item \textbf{Even in a purely dynamically typed language, contracts are an
        appealing feature for validation and correctness}. The gradual
        typing part deserved to be mentioned, but is is out of the scope of this
        paper. 
    \item \textbf{User-defined contracts are essential for validation}. Some
        contract systems (as the one of~\cite{RootCauseOfBlame}, that will be
        extensively discussed later in the paper) do not feature user-defined
        contracts.  Rather, a fixed set of built-in contracts such as
        \nickel{Number} or \nickel{String} are provided, very much like base
        types. The absence of user-defined contracts is understandable when
        using contracts as a dynamic typing discipline for correctness. But a
        system without user-defined contracts is limited in the same way as
        static typing with respect to validation.
\end{itemize}
